---
title: "Trend analysis for USFWS species status assessment for bull trout (_Salvelinus confluentus_)"
author:
  - Mark D. Scheuerell:
      email: scheuerl@uw.edu
      institute: wacfwru
      correspondence: true
institute:
  - wacfwru: U.S. Geological Survey Washington Cooperative Fish and Wildlife Research Unit, School of Aquatic and Fishery Sciences, University of Washington, Seattle, WA
output:
  pdf_document:
    highlight: haddock
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
documentclass: article
geometry: margin=1in
linestretch: 1.1
fontsize: 12pt
header-includes:
  - \usepackage{lineno}
  - \linenumbers
bibliography: "references.bib"
csl: "frontiers-in-ecology-and-the-environment.csl"
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)

## load {knitcitations}
library("knitcitations")

## clean out any existing .bib files
# cleanbib()

## set pandoc for formatting
options("citation_format" = "pandoc")
```

\vspace{4in}

This draft manuscript is distributed solely for purposes of scientific peer review. Its content is deliberative and predecisional, so it must not be disclosed or released by reviewers. Because the manuscript has not yet been approved for publication by the U.S. Geological Survey (USGS), it does not represent any official USGS finding or policy.

\newpage

# Background

The purpose of this analysis is to estimate trends in the abundance (counts) of bull trout within predefined core areas spread across Oregon, Washington, Idaho, and Montana. 

## Data

I provided a template for the data in a "tidy" format, which consisted of the following 8 fields:

* `dataset` (i.e., integer value for unique ID)
* `recovery unit` (e.g., Mid Columbia)
* `core area` (e.g., South Fork Clearwater)
* `popn/stream` (e.g., Crooked River)
* `metric` (e.g., abundance, survival)
* `method` (e.g., redd survey, electrofishing)
* `year`
* `value` (i.e., generally a count)

Data files were provided to me by data coordinators from each of the four states. They are:

* Oregon: Stephanie Gunckel (ODFW)
* Washington: Marie Winkowski (WDFW)
* Idaho: Brett Bowersox (IDFG)
* Montana: Dan Brewer (USFWS), Clint Muhlfeld (USGS), Tim Cline (USGS)

The data coordinators also provided me with some metadata indicating which of the data specific to a location were generally for adults versus juveniles.

# Modeling framework

## Population model

I fit discrete-time versions of exponential models for population growth (decline), such that the abundance of bull trout $(N)$ is a function of the initial population size $N_0$, time $(t)$, the population growth rate $(u)$, and a time-dependent stochastic effect of the environment $(w)$. Specifically, in continuous time the model is

\begin{equation}
N(t) = N_0 \exp(u) \exp(w t).
\end{equation}

In discrete time, with a time step of 1 unit (e.g., a year), the model becomes

\begin{equation}
N_t = N_{t-1} \exp(u + w_t).
\end{equation}

If we take the logarithm of both sides and define $x_t = \log(N_t)$, we have

\begin{equation}
x_t = x_{t-1} + u + w_t.
\end{equation}

Further defining $w_t \sim \text{N}(0, q)$ leads us to a so-called "biased random walk" model, where $u$ is the tendency for the population to increase or decrease each time step (i.e., the bias), and $w_t$ is some unknown stochastic of the environment.

## Observation model

The data available to us rarely come from complete censuses, and instead are typically derived from partial counts. Furthermore, sampling or observation errors may occur when counting individuals or redds. Thus, we should account for these possible errors with a so-called "data model". In this case, we can assume that the data in hand are a somewhat distorted view of the "true state of nature", such that the logarithm of the observed count at time $t$ $(y_t)$ equals the true count plus or minus some error. Specifically, we can write this as

\begin{equation}
y_t = x_t + v_t
\end{equation}

where $v_t \sim \text{N}(0, r)$.

# Model fitting

All models were fit using the `{MARSS}` package `r citep(citation("MARSS"))` for the **R** computing software `r citep(citation())`.


# References

\setlength{\parindent}{-0.375in}
\setlength{\leftskip}{0.375in}
\setlength{\parskip}{8pt}
\noindent

```{r}
## create ref cited section
write.bibtex(file = "references.bib")
```


